Providing a Unified Software Environment for Canada’s National Advanced Computing Centers

### 2019-04-08 extended-abstract.tex
Corrections
Integrated Ryan's comments

The second principle is that the software environment should be decoupled from the underlying operating system. When we first designed the solution, new clusters were being deployed with the CentOS 7 operating system. However, they did not all have the same set of system packages installed. Moreover, we were interested in supporting legacy clusters
that were running CentOS 6, and virtual machines launched by users on one of several Compute Canada cloud platforms. In order to use the unified software environment on all of these systems, we needed a compatibility layer. 
The solution we deployed is Nix \cite{Nix}, although we are experimenting with Gentoo Prefix \cite{Gentoo} as a replacement. We will discuss these options in Section~\ref{sec:Compatibility_layer}.

@@

\section{Distribution mechanism}
\label{sec:Distribution_mechanism}
A shared file system such as NFS, Lustre, or GPFS, can provide access to software packages on a single cluster. However, experience in the field of high performance computing shows that parallel file systems are often challenging to operate at scale in a reliable and performant manner, especially for the high I/O operations per second caused by applications startup. In addition, a shared file system is a single point of failure for a cluster, which nearly all jobs rely on in order to run. An alternative for a single cluster is to install binary packages on each node, using a mechanism such as RPM. However, this can result in divergent configurations across nodes, may be limited by available space on the local disk, and requires an additional configuration management layer to keep the packages synchronized.

@@

\subsection{CVMFS structure}
\label{sub:CVMFS_structure}
CVMFS is designed similarly to the Network Time Protocol stratum model. There is one source of truth called the stratum 0, where software is injected.  The stratum 0 is never accessed directly by clients. If it fails, the only immediate consequence is that injection of new files is not possible until it is back online.

Snapshots of the stratum 0 are kept for a while, such that it is possible to roll back to a previous version if needed. Files in the repository are compressed, and their hash is computed to save space by deduplicating objects. Hashes also allow CVMFS clients to quickly verify and invalidate their local caches when content is changed.

The stratum 0 is replicated to multiple strata 1 servers, which serve content to clients and provide redundancy and load balancing. Files are served to the clusters via HTTP, and each cluster has local caching proxy servers to reduce the load placed on the stratum 1 servers by its worker nodes mitigating the impact of a wide area network outage. The closest layer of cache can be configured to use a local disk on the compute nodes, or a shared file system in a diskless environment.

>>>>

The second principle is that the software environment should be decoupled from the underlying operating system. When we first designed the solution, new clusters were being deployed with the CentOS 7 operating system. However, they did not all have the same set of system packages installed. Moreover, we were interested in supporting legacy clusters that were running CentOS 6, and virtual machines launched by users on several Compute Canada cloud platforms. In order to use the unified software environment on all of these systems, we needed a compatibility layer. 
The solution we deployed is Nix \cite{Nix}, although we are experimenting with Gentoo Prefix \cite{Gentoo} as a replacement. We will discuss these options in Section~\ref{sec:Compatibility_layer}.

@@

\section{Distribution mechanism}
\label{sec:Distribution_mechanism}
A shared file system such as NFS, Lustre, or GPFS can provide access to software packages on a single cluster. However, experience in the field of high performance computing shows that parallel file systems are often challenging to operate at scale in a reliable and performant manner, especially for the high I/O operations per second caused by applications starting. In addition, a shared file system is a single point of failure for a cluster, which nearly all jobs rely on in order to run. An alternative for a single cluster is to install binary packages on each node, using a mechanism such as RPM. However, this can result in divergent configurations across nodes, may be limited by available space on the local disks, and requires an additional configuration management layer to keep the packages synchronized.

@@

\subsection{CVMFS architecture}
\label{sub:CVMFS_structure}
CVMFS is designed similarly to the Network Time Protocol stratum model. There is one source of truth, the stratum 0 server, where software is injected.  The stratum 0 is never accessed directly by clients. If it fails, the only immediate consequence is that injection of new files is not possible until it is back online.

Snapshots of the software repositories on the stratum 0 are kept for a while, making it possible to roll back to a previous version if needed. Internally, CVMFS uses a content-addressable storage scheme, wherein files exceeding a certain size are broken into smaller objects, and each object is hashed, compressed, and stored in a location defined by its hash, resulting in natural deduplication. Hashes also allow CVMFS clients to quickly verify the integrity of content, and invalidate their local caches when content is changed, by means of a Merkle tree.

The stratum 0 is replicated to multiple stratum 1 servers, which serve content to clients and provide redundancy and load balancing. Files are served to the clusters via HTTP, and each cluster has local caching proxy servers to reduce the load placed on the stratum 1 servers by its worker nodes, which also mitigates the impact of a wide area network outage. The closest layer of cache can be configured to use a local disk on the compute nodes, or a shared file system in a diskless environment. Clients can also be configured to use two levels of cache together in a tiered manner, such as a solid state drive and a hard drive, or a hard drive and a shared file system.

### 2019-04-08 paper.tex
Integrated Ryan's comments

\section{Distribution mechanism}
\label{sec:Distribution_mechanism}
A shared file system such as NFS, Lustre, or GPFS can provide access to software packages on a single cluster. However, experience in the field of high performance computing shows that parallel file systems are often challenging to operate at scale in a reliable and performant manner, especially for the high I/O operations per second caused by application startups. In addition, a shared file system is a single point of failure for a cluster, which nearly all jobs rely on in order to run. An alternative for a single cluster is to install binary packages on each node, using a mechanism such as RPM. However, this can result in divergent configurations across nodes, may be limited by available space on the local disk, and requires an additional configuration management layer to keep the packages synchronized.

For a national platform spanning multiple clusters, latency precludes a parallel file system from consideration, and the synchronization problem is compounded. The clusters could be synchronized by pulling or pushing changes with a tool such as {\it rsync}, but this would not be scalable; even a tool as efficient as {\it rsync} takes several hours to synchronize over 40 million files and directories. Moreover, this method of copying files is not transactional or atomic, so an additional indirection layer would be required to provide a complete solution.

We therefore chose to use CERN Virtual Machine File System (CVMFS) \cite{CVMFS}. CVMFS is a file system designed to distribute software packages to clusters at a global scale with no single point of failure. Clients access files by pulling content on demand, using multiple levels of caching, to minimize network traffic and storage use and ensure responsive and scalable performance.

We will describe the generic structure of a CVMFS deployment in Section~\ref{sub:CVMFS_structure}, and describe the deployment specific to Compute Canada in Section~\ref{sub:CVMFS_in_CC}

\subsection{CVMFS structure}
\label{sub:CVMFS_structure}
CVMFS is designed similarly to the Network Time Protocol stratum model, as illustrated in Figure~\ref{fig:CVMFS_structure}. There is one source of truth called the stratum 0, where software is injected.  The stratum 0 is never accessed directly by clients. If it fails, the only immediate consequence is that injection of new files is not possible until it is back online.

Snapshots of the stratum 0 are kept for a while, making it possible to roll back to a previous version if needed. Files in the repository are compressed, and their hash is computed to save space by deduplicating objects. Hashes also allow CVMFS clients to quickly verify and invalidate their local caches when content is changed.

The stratum 0 is replicated to multiple strata 1 servers, which serve content to clients and provide redundancy and load balancing. Files are served to the clusters via HTTP, and each cluster has local caching proxy servers to reduce the load placed on the stratum 1 servers by its worker nodes, and mitigate the impact of a wide area network outage. The closest layer of cache can be configured to use a local disk on the compute nodes, or a shared file system in a diskless environment.

\begin{figure}
  \includegraphics[width=0.45\textwidth]{CVMFS-structure.pdf}
  \caption{Diagram of the CVMFS structure. Arrows represent the flow of data. Data is pushed into the stratum-0, but is then fetched by the following layers.}
  \label{fig:CVMFS_structure}
\end{figure}

\subsection{Deployment of CVMFS in Compute Canada}
\label{sub:CVMFS_in_CC}
For Compute Canada's needs, we have deployed three CVMFS repositories, each composed of a stratum 0 and one or multiple strata 1. This allows us to follow a workflow that includes testing before exposing software packages to user, as illustrated in Fig~\ref{fig:Workflow}.

First, we have a development repository. This repository is used to deploy software from our build infrastructure onto the clusters without exposing it to end users. This allows us test newly installed software packages on our clusters without impacting production. This minimizes the odds of a faulty package being pushed to production. This is especially important when a bug is found in one of the software packages and we need to update it. This also happens for example if we need to make minor changes to an MPI implementation in order to correctly support a new cluster with a different networking fabric.

We then have a production repository. By design, software packages are always copied onto the production repository from the development repository, and never from the build infrastructure directly. This workflow also minimizes the odds of a faulty software making its way to the end users. This repository holds all of the software environment which is visible to the end users on all of our clusters. Once a piece of software is deployed onto this repository, it becomes visible on all clusters within 30 minutes.

Both repositories above are available publicly, and can be mounted on any cluster or virtual machine in the cloud. We also have a third repository which is used to host software packages that are restricted to Compute Canada due to license limitations. We work with software vendors and developers in order to get the authorization to install licensed and commercial software packages in this repository. The most common licensing model that we use is a BYOL (bring your own license) model, in which we provide the software package, but end users provide a license server to connect to. Another model that is used is to control access to a software package via POSIX groups. By default, CVMFS clients do not honour POSIX group and filesystem permissions, but they can be configured to do so. Since this kind of restriction requires a specific LDAP and a specific configuration, this repository is restricted and can only be made available to clusters that are run by Compute Canada teams.

One final use case for CVMFS is to hold and distribute datasets. However, default CVMFS configuration is optimized for a large number of small files, rather than a small number of large files. This means that a special configuration must be done for this repository. At the time of writing this paper, this use case is still experimental within Compute Canada, but other groups are known to use CVMFS in this way.

@@

The second layer is what we call our compatibility layer. It contains almost all packages that one would typically install with the Linux distribution's package manager, such as {\it yum} or {\it apt-get}.

@@

The installation path of each software package installed through Nix resides in a path called the Nix store. This stores contains a single directory for each version of each software package installed through Nix. These directories are named after a hash calculated from the installation recipe and its dependencies. This creates a very large number of subdirectories, many of which contain almost identical files.

In addition to the store, Nix creates directories that are called profiles. These profile directories contain a farm of symbolic links that point to files in the Nix store. For example, it will contain a {\it bin} directory, which will contain symbolic links to typical operating system commands such as {\it ls} or {\it gcc}, which will reside in different paths in the store. It will also contain {\it lib} and {\it include} directories which will contain symbolic links to libraries and header files located in the store.

Profiles are the paths that are exposed to the end users, which is done by adding paths to their {\it PATH}, {\it CPATH} and other similar environment variables. Each profile ever created is kept, but users are only exposed to the most recent one. Each profile represents the complete state of the user-exposed environment at a given time in the past.

@@

This works well until a software exposes paths to the Nix store directly, instead of the paths to Nix profiles. We call this a Nix store {\it leak}. This happens for example with Python when using virtual environments. Virtual environments in Python copy the {\it python} executable to a secondary location, outside of Nix. However, all binaries installed by Nix use {\it RPATH} for shared object resolutions in the store directly. This means that copying a binary outside of Nix creates a dependency between a binary that now resides outside of the store, and the store itself. Because Nix is unaware of that dependency, any update of one of the dependencies followed by garbage collection or selective deployment of files will end up breaking this binary. We have encountered this issue with Python, Perl and Qt, which we have therefore moved outside of Nix and installed them through EasyBuild instead.

Another issue that we have encountered is that many Nix recipes call for installing sub-packages of a package in separate directories. For example, for Qt, {\it qmake} and {\it OTHER COMMAND} reside in different directories in the Nix store. This sometimes cause issues when installing a dependent package which assumes that its dependencies are installed in a more traditional fashion, with all of the binaries in the same {\it bin} folder, alongside with headers in the {\it include} folder, and libraries in the {\it lib} folder. This has mostly been an issue for Qt.

\subsection{Gentoo Prefix}
\label{sub:Gentoo_Prefix}
Gentoo Prefix~\cite{Gentoo} is an alternative solution to provide a compatibility layer. It allows installation of most packages that form a typical Linux operating system in a non-default location. Contrary to Nix however, packages are installed in a standard directory structure ({\it bin}, {\it lib}, {\it include}, {\it share}). This is helpful when updates are required, and minimizes problems with applications that expect a standard directory structure. At the time of this writing, we are still experimenting with Gentoo Prefix, and it is not used in production.

@@

\subsection{Compiled software packages}
\label{sub:Compiled_software_packages}
Scientific applications constitute the majority of our software stack. Since their performance is critical, they can not be installed in the compatibility layer, as they need to be optimized for specific CPU instructions. To install them, we use the EasyBuild~\cite{EasyBuild2012,EasyBuild2014,EasyBuild2016} software installation tool.

EasyBuild originated from the University of Ghent, but now has over 200 contributors from around the world. The framework is based on scripted recipes (in Python) to install packages, usually from source code. It is designed to create optimized binaries, and eases recompilation of multiple versions of a given application using many combinations of compilers or libraries. EasyBuild is composed of three parts: the framework, easyblocks, and easyconfig files (also known as easyconfigs).

The framework contains all of the core code which drives the logic of installing a software. It drives all of the steps necessary to install a software: parsing the configuration options, downloading source files, patching them, configuring, building and installing the software, as well as performing post-installation steps such as sanity checks, generating a module or running custom commands.

Easyblocks are Python classes that define the steps required to install a given software package. There are generic easyblocks, such as {\it ConfigureMake} or {\it CMakeMake}, as well as custom easyblocks for packages that have more complicated installation procedures, such as OpenFOAM or PETSc.

Easyconfig files parametrize the execution of the code from the easyblocks. They define the software name and version, the URL from which the source files can be downloaded as well as options to be used for the various steps.

@@

Toolchains are a core concept in EasyBuild. They represent a set or subset of a compiler, an MPI implementation, mathematical libraries and CUDA. Having the ability to generate multiple builds of a given software simply by providing a command line argument to change the toolchain has proven invaluable to our workflow, and has allowed us to recompile hundreds of packages in matter of days when deploying a new version of compiler or a cluster with a new CPU architecture.

EasyBuild also automatically creates module files~\cite{Modules1991,Modules1996,Lmod}. The framework supports multiple standard and custom naming schemes. We use a version of a hierarchical naming scheme, in which modules are organized in a tree structure, hiding those that are incompatible with currently loaded modules. This limits the list of visible modules which could otherwise be very extensive. This can however throw off some users, since they don’t automatically see all available modules. Lmod~\cite{Lmod} alleviates this issue by providing a separate command to crawl the full module tree and instructs users on how to load the module they want.

@@

\subsection{Packages for scripting languages}
\label{sub:Packages_for_scripting_languages}
For scripting languages that support installing packages, such as Python, R or Perl, we leverage the existing package manager of that language, rather than installing them as modules. In the case of Python, we are building Python wheels~\cite{Wheels}, linking against our software stack. Wheels are binary packages that can be installed in a matter of seconds, similar to CentOS RPM packages. Since they are built by our team, and linked against our optimized software stack, we ensure that they are easy to install and offer good performance.

\subsection{Licensed and restricted software packages}
\label{sub:Licensed_software_packages}
Compute Canada also has agreements with a number of software vendors or developers of applications that require a license. In these cases, we install the package using the same tools, but it is deployed on a separate CVMFS repository which is only accessible to Compute Canada clusters.

\subsection{Documentation of supported packages}
\label{sub:Documentation}
Every software package that is installed through EasyBuild get automatically listed on our documentation website, at \\
\url{https://docs.computecanada.ca/wiki/Available_software}
A similar list of available Python wheels is automatically generated at \\
\url{https://docs.computecanada.ca/wiki/Available_wheels}

\subsection{Logging of module usage}
\label{sub:Logging}
Our module system will automatically log (through {\it logger}) each module that is loaded. The information is logged in an easily parsable list of key-value pairs, which can then be aggregated into your favorite monitoring tool such as Elastic Search.

\section{Making the software environment available on your cluster or computer}
\label{sec:Making_it_available}
Making our software environment available on a single computer is as simple as installing the CVMFS client and configuring repositories manually or installing our RPM
\textbf{TODO: Document our RPM}

After CVMFS is configured, one can simply execute the command below in order to obtain Compute Canada's software environment \\
\texttt{source /cvmfs/soft.computecanada.ca/config/profile/bash} \\

By default, our default set of modules will try to detect the most recent set of instructions that is supported by the CPU amongst {\it SSE3}, {\it AVX}, {\it AVX2} and {\it AVX512}. It will also try to detect whether the hardware has a {\it Mellanox InfiniBand}, {\it Intel OmniPath}, or an Ethernet network fabric. If these are not detected correctly, or if you have a cluster made up of mixed hardware, you can overwrite these defaults by defining the {\verb!RSNT_ARCH!} or {\verb!RSNT_INTERCONNECT!} environment variables before executing the command above.

Note that the first time a command is ran, it will take some longer time to start the command, as the CVMFS client is fetching the files that are needed to run the command from our stratum 1 servers. Following executions of the same command will however be fast as files will have been cached locally.

Also note that the command above will \textbf{not} import our environment for any user ID lower than 1000, which are typically privileged users on a Linux system. In order to force the command above to execute nevertheless, you can create a file named \verb!.force_cc_cvmfs! in the user's \verb!$HOME! directory.

Making our software environment available on a cluster requires the additional step of configuring one or multiple HTTP proxies. This is required to avoid overloading the stratum-1 server by having a large number of nodes all fetching the same files at the same time. If you are interested in testing or making our software environment available to your users on your cluster, we ask that you first contact us. Note that this software stack comes with absolutely no warranty.

\section{Other benefits and conclusion}
\label{sec:Conclusion}
We have built a software environment that is portable, is optimized for multiple CPU architectures and network interconnect, and can be deployed in a matter of minutes on a single node, and barely longer for a whole cluster. The design of the infrastructure guarantees that the software environment seen by our end users is the same across all of our clusters, and will not diverge over time. This makes for a better user experience, as it becomes much easier to migrate between two clusters. It also reduces the workload on our staff by unifying the installation process for all clusters, and makes supporting our users easier because all clusters work the same way.

Because of the portability of our software stack and the distribution mechanism, it can be used on any cluster, or any virtual machine on a cloud. For example, for training events, we have been using it to provision small replicas of Compute Canada clusters in short-lived virtual clusters deployed on our OpenStack platform. We have also used this portable software environment to expose a cluster environment directly through a browser, via JupyterHub. It can also help support infrastructures owned by various institutions.

@@

%
% The next two lines define the bibliography style to be used, and the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{biblio}

>>>>

\section{Distribution mechanism}
\label{sec:Distribution_mechanism}
A shared file system such as NFS, Lustre, or GPFS can provide access to software packages on a single cluster. However, experience in the field of high performance computing shows that parallel file systems are often challenging to operate at scale in a reliable and performant manner, especially for the high rate of I/O operations per second caused by applications starting. In addition, a shared file system is a single point of failure for a cluster, which nearly all jobs rely on in order to run. An alternative for a single cluster is to install binary packages on each node, using a mechanism such as RPM. However, this can result in divergent configurations across nodes, may be limited by available space on the local disks, and requires an additional configuration management layer to keep the packages synchronized.

For a national platform spanning multiple clusters, latency precludes a parallel file system from consideration, and the synchronization problem is compounded. The clusters could be synchronized by pulling or pushing changes with a tool such as \texttt{rsync}, but this would not be scalable; even a tool as efficient as \texttt{rsync} takes several hours to synchronize over 40 million files and directories. Moreover, this method of copying files is not transactional or atomic, so an additional indirection layer would be required to provide a complete solution.

We therefore chose to use CERN Virtual Machine File System (CVMFS) \cite{CVMFS,CVMFS2012}. CVMFS is a file system designed to distribute software packages to clusters at a global scale with no single point of failure. Clients access files by pulling content on demand, using multiple levels of caching, to minimize network traffic and storage use and ensure responsive and scalable performance.

We will describe the general architecture of a CVMFS deployment in Section~\ref{sub:CVMFS_structure}, and describe the deployment specific to Compute Canada in Section~\ref{sub:CVMFS_in_CC}

\subsection{CVMFS architecture}
\label{sub:CVMFS_structure}
CVMFS is designed similarly to the Network Time Protocol stratum model, as illustrated in Figure~\ref{fig:CVMFS_structure}. There is one source of truth, the stratum 0 server, where software is injected.  The stratum 0 is never accessed directly by clients. If it fails, the only immediate consequence is that injection of new files is not possible until it is back online.

Snapshots of the software repositories on the stratum 0 are kept for a while, making it possible to roll back to a previous version if needed. Internally, CVMFS uses a content-addressable storage scheme, wherein files exceeding a certain size are broken into smaller objects, and each object is hashed, compressed, and stored in a location defined by its hash, resulting in automatic deduplication. Hashes also allow CVMFS clients to quickly verify the integrity of content received over the network, and invalidate their local caches when content is changed, by means of a Merkle tree.

Data is replicated from the stratum 0 to multiple stratum 1 servers, which in turn serve content to clients via HTTP and provide redundancy and load balancing. Each cluster has local caching proxy servers to reduce the load placed on the stratum 1 servers by its compute nodes, accelerate access to any content missing from the local cache of a compute node, and mitigate the impact of a wide area network outage. The closest layer of cache can be configured to use a local disk on a node, or a shared file system in a diskless environment. Clients can also be configured to use two levels of cache together in a tiered manner, such as a solid state drive and a hard drive, or a hard drive and a shared file system.

\begin{figure}
  \includegraphics[width=0.45\textwidth]{CVMFS-structure.pdf}
  \caption{Diagram of the architecture of CVMFS. Arrows represent the flow of data. Data is pushed to the stratum 0, but is then fetched by the following layers.}
  \label{fig:CVMFS_structure}
\end{figure}

\subsection{Deployment of CVMFS in Compute Canada}
\label{sub:CVMFS_in_CC}
For Compute Canada's needs, we deployed two CVMFS server stacks, one for development and one for production, each of which is composed of a stratum 0 server and one or more stratum 1 servers. This allows us to follow a workflow that includes testing software packages before exposing them to users, as illustrated in Fig~\ref{fig:Workflow}. In total, three software repositories are used.

First, we have a development repository, hosted on the development servers. This repository is used to deploy software from our build infrastructure to the clusters without exposing it to end users. This allows us to test newly installed software packages on our clusters without impacting production, minimizing the risk of a faulty package causing problems for users. This is especially important, for example, when testing updates to software packages to fix bugs, or to make minor changes to an MPI implementation in order to correctly support a new cluster with a different networking fabric.

We also have a production repository, hosted on the production servers. By design, software packages are always copied to the production repository from the development repository, and never directly from the build infrastructure. This step also minimizes the risk of faulty software affecting end users. This repository holds all of the software environment which is visible to the end users on all of our clusters. Once software is published to this repository, it becomes visible on all clusters within about 10 minutes.

Both repositories above are available publicly, and can be mounted on any cluster or virtual machine in the cloud. There is a third repository, also hosted on the production servers, which provides software packages that are restricted to Compute Canada due to license limitations. We work with software vendors and developers in order to get the authorization to install licensed and commercial software packages in this repository. The most common licensing model that we use is a BYOL (bring your own license) model, in which we provide the software package, but end users provide a license server to connect to. Another model is to control access to a software package via POSIX groups. By default, the repository content on a CVMFS client belongs to a \texttt{cvmfs} service account and group, but this can be changed so that the original ownership metadata as published on the stratum 0 is used, thereby only allowing access to members of authorized POSIX groups. Since this relies on a specific LDAP domain and client configuration, access to this repository is restricted by network ACLs on the stratum 1 servers, and can only be granted to clusters that are run by Compute Canada teams.

There is a fourth CVMFS repository that distributes configuration files instead of software. This special configuration repository resembles \texttt{/etc/cvmfs/} and provides information (such as URLs and public keys) required by CVMFS clients in order to access all {\it other} repositories. The benefit of this approach is that most CVMFS configuration can be centrally managed, and updates can be immediately pushed to all clients in order to set up a new repository or commission or decommission a stratum 1 server, rather than having to wait for all clients across the country to be updated.

One final use case for CVMFS is to distribute datasets. By default, CVMFS  is optimized for a large number of small files, rather than a small number of large files, so different  configurations and access methods are needed for data repositories. At the time of writing this paper, this is used by some of our research groups in genomics, but not directly by our staff.

@@

The second layer is what we call our compatibility layer. It contains almost all packages that one would typically install with the Linux distribution's package manager, such as \texttt{yum} or \texttt{apt-get}.

@@

The installation path of each software package installed through Nix resides in a path called the Nix store. This store contains a single directory for each version of each software package installed through Nix. These directories are named after a hash calculated from the installation recipe and its dependencies. This creates a very large number of subdirectories, many of which contain almost identical files.

In addition to the store, Nix creates directories that are called profiles. These profile directories contain a forest of symbolic links that point to files in the Nix store. For example, it contains a \texttt{bin/} directory, which contains symbolic links to typical operating system commands such as \texttt{ls} and \texttt{gcc}, which reside in different paths in the store. A profile directory also contains \texttt{lib/} and \texttt{include/} directories which contain symbolic links to libraries and header files located in the store.

Profiles are the paths that are exposed to the end users, which is done by adding paths to their \texttt{PATH}, \texttt{CPATH} and other similar environment variables. Each profile ever created is kept, but users are only exposed to the most recent one. Each profile represents the complete state of the user-exposed environment at a given time.

@@

This works well until a software package exposes paths to the Nix store directly, instead of the paths to Nix profiles. We call this a Nix store leak. This happens for example with Python when using virtual environments. Virtual environments in Python copy the \texttt{python} executable to a secondary location, outside of Nix. However, all binaries installed by Nix use \texttt{RPATH} for shared object resolutions in the store directly. This means that copying a binary outside of Nix creates a dependency between a binary that now resides outside of the store, and the store itself. Because Nix is unaware of that dependency, any update of one of the dependencies followed by garbage collection or selective deployment of files will end up breaking this binary. We have encountered this issue with Python, Perl and Qt, which we have therefore moved outside of Nix and installed through EasyBuild instead.

Also, we have found that many Nix recipes call for installing sub-packages of a package in separate directories. For example, for Qt, \texttt{qmake} and \texttt{qtpaths} or \texttt{qtdiag} reside in different directories in the Nix store. This sometimes causes issues when installing a dependent package which assumes that its dependencies are installed in a more traditional fashion, with all of the binaries in the same \texttt{bin/} directory, alongside headers in the \texttt{include/} directory and libraries in the \texttt{lib/} directory.

\subsection{Gentoo Prefix}
\label{sub:Gentoo_Prefix}
Gentoo Prefix~\cite{Gentoo} is an alternative solution to provide a compatibility layer. It allows installation of most packages that form a typical Linux operating system, in a non-default location. Contrary to Nix however, packages are installed in a standard directory structure (\texttt{bin/}, \texttt{lib/}, \texttt{include/}, \texttt{share/}). This is helpful when updates are required, and minimizes problems with applications that expect a standard directory structure. At the time of this writing, we are still experimenting with Gentoo Prefix, and it is not used in production.

@@

\subsection{Compiled software packages}
\label{sub:Compiled_software_packages}
Scientific applications constitute the majority of our software stack. Since their performance is critical, they can not be installed in the compatibility layer, as they need to be optimized for specific CPU instruction sets. To install them, we use the EasyBuild~\cite{EasyBuild2012,EasyBuild2014,EasyBuild2016} software installation tool.

EasyBuild originated from the University of Ghent, but now has over 200 contributors from around the world. The framework is based on scripted recipes (in Python) to install packages, usually from source code. It is designed to create optimized binaries, and eases recompilation of multiple versions of a given application using many combinations of compilers or libraries. EasyBuild is composed of three parts: the framework, easyblocks, and easyconfig files (also known as ``easyconfigs'').

The framework contains all of the core code which drives the logic of installing a software package. It includes the following steps: parsing the configuration options, downloading source files, patching them, configuring, building and installing the software, as well as performing post-installation steps such as sanity checks, generating a module or running custom commands.

Easyblocks are Python classes that define the steps required to install a given software package. There are generic easyblocks, such as \texttt{ConfigureMake} or \texttt{CMakeMake}, as well as custom easyblocks for packages that have more complicated installation procedures, such as OpenFOAM or PETSc.

Easyconfig files parameterize the execution of the code from the easyblocks. They define the software name and version, the URL from which the source files can be downloaded, and options to be used for the various steps.

@@

Toolchains are a core concept in EasyBuild. They represent a set or subset of a compiler, an MPI implementation, mathematical libraries and CUDA. Having the ability to generate multiple builds of a given software simply by providing a command line argument to change the toolchain has proven invaluable to our workflow, and has allowed us to recompile hundreds of packages in a matter of days when deploying a new version of compiler or a cluster with a new CPU architecture.

EasyBuild also automatically creates module files~\cite{Modules1991,Modules1996,Lmod}. The framework supports multiple standard and custom naming schemes. We use a version of a hierarchical naming scheme, in which modules are organized in a tree structure, hiding those that are incompatible with currently loaded modules. This limits the list of visible modules which could otherwise be very extensive. This can however confuse some users, since they don’t automatically see all available modules. Lmod~\cite{Lmod} alleviates this issue by providing a separate command to crawl the full module tree and instructs users on how to load the module they want.

@@

\subsection{Packages for scripting languages}
\label{sub:Packages_for_scripting_languages}
For scripting languages that support installing packages, such as Python, R or Perl, we leverage the existing package manager of that language, rather than installing them as modules. In the case of Python, we build Python wheels~\cite{Wheels}, linked against our software stack. Wheels are binary packages that can be installed in a matter of seconds, similar to RPM packages. Since they are built by our team, and linked against our optimized software stack, we ensure that they are easy to install and offer good performance.

\subsection{Documentation of supported packages}
\label{sub:Documentation}
Every software package that is installed through EasyBuild gets automatically listed on our documentation website:
\begin{center}
    \url{https://docs.computecanada.ca/wiki/Available_software}
\end{center}
Similarly, a list of available Python wheels is automatically generated:
\begin{center}
\url{https://docs.computecanada.ca/wiki/Available_wheels}    
\end{center}

\subsection{Logging of module usage}
\label{sub:Logging}
Our module system will automatically log (through \texttt{logger}) each module that is loaded. The information is logged in an easily parsable list of key-value pairs, which can then be aggregated into your favorite monitoring tool, such as Elastic Search.

\section{Making the software environment available on your system}
\label{sec:Making_it_available}
Making our software environment available on a single computer is as simple as installing the CVMFS client and configuring repositories manually or installing our RPM
\textbf{TODO: Document our RPM}

After CVMFS is configured, one can simply execute this command in order to set up Compute Canada's software environment:

\begin{center}
\texttt{source \textbackslash \linebreak
/cvmfs/soft.computecanada.ca/config/profile/bash.sh}
\end{center}

The default set of modules will automatically try to detect the most recent set of instructions that is supported by the CPU amongst  SSE3, AVX, AVX2 and AVX512. It will also try to detect whether the hardware has a Mellanox InfiniBand, Intel OmniPath, or Ethernet network fabric. If these are not detected correctly, or if you have a cluster made up of mixed hardware, you can overwrite these defaults by defining the \texttt{RSNT\_ARCH} or \texttt{RSNT\_INTERCONNECT} environment variables before executing the command above.

Note that the first time any software or data is accessed, it will take longer to load, as the CVMFS client is fetching the content from our stratum 1 servers. However, subsequent access will be much faster once the content is cached locally. 
Also note that the command above will {\it not} import our environment for any user IDs lower than 1000, which are typically privileged users on a Linux system. In order to force the command above to execute nevertheless, you can create a file named \texttt{.force\_cc\_cvmfs} in the user's \texttt{\$HOME} directory. Conversely, a \texttt{\$HOME/.skip\_cc\_cvmfs} file can be created to exclude a user from setting up the software environment. This is recommended for system administrators so their environment does not depend on CVMFS, especially if they are doing operations which may interrupt access to CVMFS, such as updating or reconfiguring it.

If you plan to use our CVMFS repositories on a large number of nodes such as a cluster, we require that you deploy one or more caching HTTP proxies, to avoid overloading the stratum 1 servers. If you are interested in testing or making our software environment available to your users on your cluster, we ask that you first contact us by writing to 
\begin{center}
\url{rsnt-software@computecanada.ca}    
\end{center}
Note that this software stack comes with absolutely no warranty.

\section{Other benefits and conclusion}
\label{sec:Conclusion}
We have built a software environment that is portable, is optimized for multiple CPU architectures and network interconnects, and can be deployed in a matter of minutes on a single node, and not much longer for a whole cluster. The design of the infrastructure guarantees that the software environment seen by our end users is the same across all of our clusters, and will not diverge over time. This makes for a better user experience, as it becomes much easier to migrate between two clusters. It also reduces the workload on our staff by unifying the installation process for all clusters, and makes supporting our users easier because all clusters work the same way.

Because of the portability of our software stack and the distribution mechanism, it can be used on any cluster, or any virtual machine on a cloud. For example, for training events, we have been using it to provision small replicas of Compute Canada clusters in short-lived virtual clusters deployed on our OpenStack platform. We have also used this portable software environment to expose a cluster environment directly through a browser, via JupyterHub. It can also help support infrastructures owned by various institutions. This is the case of many universities in Canada that used to host a cluster which is now used by the institution only, or universities that deploy new local infrastructure and don't have the human resources to build a large software stack.

@@

% The next two lines define the bibliography style to be used, and the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{biblio}
